---
title: "Appendix to _Measuring Children's Early Vocabulary in Low-Resource Languages Using a Swadesh-style Word List_"
author: "[redacted for anonymous review]"
date: "May 7, 2025"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=F, message=F)

library(png)
library(grid)
#library(xtable)
require(mirt)
require(tidyverse)
require(ggpubr)
require(tidyboot)
require(here)
library(RColorBrewer)
library(gplots)
library(kableExtra)

html_table_width <- function(kable_output, width){
  width_html <- paste0(paste0('<col width="', width, '">'), collapse = "\n")
  sub("<table>", paste0("<table>\n", width_html), kable_output)
} 

source(here("scripts/04-cross-ling-comparison-helpers.R"))

# production fits
load(here("data/intermediates/xldf_clean_allforms.Rdata")) 
# xldf_clean, prod_pars
# 42 languages, 28541 parameters

xldf_clean <- xldf_clean |> 
  mutate(uni_lemma = ifelse(uni_lemma=="glue", "glue (object)", uni_lemma)) |>
  mutate(d = -d) |> # easiness -> difficulty
  mutate(category = case_when(category=="directions" ~ "locations", # Mandarin
                              category=="sounds2" ~ "sounds", # Japanese - oink
                              category=="conversations" ~ "games_routines", # Japanese - ok, no, hello
                              category=="others" & uni_lemma=="also" ~ "quantifiers", # Japanese
                              category=="others" & uni_lemma=="already" ~ "time_words", # Japanese
                              TRUE ~ category))

prod_pars <- prod_pars |> 
  mutate(uni_lemma = ifelse(uni_lemma=="glue", "glue (object)", uni_lemma)) |>
  mutate(d = -d) |> # easiness -> difficulty
  mutate(semantic_category = case_when(semantic_category=="directions" ~ "locations",
                              semantic_category=="sounds2" ~ "sounds",
                              semantic_category=="conversations" ~ "games_routines",
                              semantic_category=="others" & uni_lemma=="also" ~ "quantifiers",
                              semantic_category=="others" & uni_lemma=="already" ~ "time_words",
                              TRUE ~ semantic_category))

load(file=here("data/intermediates/generalization_results.rds"))
# gen_res_sum, gen_res

model_stats <- readRDS(here("data/intermediates/model_stats.rds"))

gen_langs <- model_stats |> filter(participants < 300) |> pull(language)

xldf_lowd <- xldf_clean %>% filter(is.element(language, gen_langs)) # 6324
xldf_train <- xldf_clean %>% filter(!is.element(language, gen_langs)) # 22217


olap <- get_uni_lemma_overlap(xldf_clean)
```

## Unilemma Operationalisation and Updating

Unilemmas are rough cross-linguistic conceptual mappings for words. 
For example, "dog" (English), "perro" (Spanish), and "chien" (French) all seem like they describe the same concept, _dog_. 
We thus have _dog_ as the unilemma for this concept, allowing mapping between different forms in different languages.

Decisions about which words correspond to which unilemmas are necessarily somewhat arbitrary. 
Our design intention for unilemmas was to provide reasonable mappings, lumping together many meanings rather than splitting them, allowing us to build predictive models of cross-linguistic similarities and differences in acquisition.
If we split the conceptual world too finely, it would be difficult to understand how similar the information sources are; if we split too coarsely, we can at least discover that languages differ by the differences in predictors.

However, we also preserve distinctions for concepts that are _prima facie_ similar, but that play very different cultural roles.
For example, although "pan" (Spanish) and "tortilla" (Spanish) are relatively similar, they plan different cultural roles in Mexican Spanish, and hence we err towards splitting these into separate categories.

There remained other issues regarding unilemma definitions, including cases of polysemy, handling syntactic categories, and pronouns; in each case, we made decisions based on our best guess of the most basic or common interpretation in the child language context.
Further details can be found in our unilemma policy document, available at [https://github.com/langcog/update_unilemmas/blob/main/unilemma_policy.md](https://github.com/langcog/update_unilemmas/blob/main/unilemma_policy.md).

We then recruited native or advanced proficient speakers of all the languages in Wordbank to help verify and correct unilemmas for the CDI forms. 
Subsequently, two of the authors (GK, AWMT) reviewed the suggestions and made decisions about conflicts, and standardized unilemmas into a canonical form across all instruments.
More details about the recent update can be found at [https://github.com/langcog/update_unilemmas](https://github.com/langcog/update_unilemmas).

## Unilemma Overlap on CDI Forms

Number of overlapping concepts on each pair of CDI forms.
American English has the highest average degree of overlap (458 unilemmas), which may be driven in large part by it being the oldest form, from which many others were adapted.
The form with the lowest average degree of overlap is Chilean Spanish, with only 314 unilemmas shared with other forms, on average.
The median of the average overlap per language is 393 unilemmas.

```{r, fig.width=10, fig.height=10}
# sort(rowMeans(olap))
# show heatmap of overlap between each CDI:WS
heatmap.2(olap, dendrogram = "none", margins=c(10,10))
```


## Word Difficulty by Semantic Category and Language

```{r semantic-cats-by-lang, fig.width=10, fig.height=18, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Mean difficulty of CDI words by semantic category and language. Bars represent bootstrapped 95\\% confidence intervals."}
# conversations (JP) -> "games_routines" ?
# sounds2 (JP) -> "sounds" ?
# prepositional_pronouns -> ... "connecting_words" ?

# ToDo: remake with just Swadesh list
prod_cat <- xldf_clean %>% group_by(language, category) %>%
  filter(!is.na(d), !is.na(category),
         !is.element(category, c("final_particles", "numbers", "articles", "other",
                                 "prepositional_pronouns",
                                 "descriptive_words (adverbs)", "states", "unknown",
                                 "verb_endings", "verb_modifiers",
                                 "classifiers", "locations_quantities_adverbs"))) %>%
  tidyboot::tidyboot_mean(d, na.rm=T)

prod_cat %>% ggplot(aes(x=reorder(language, mean), y=mean)) + geom_point(alpha=.7) +
  geom_linerange(aes(ymin=ci_lower, ymax=ci_upper), alpha=.7) +
  facet_wrap(. ~ category) + coord_flip() + 
  theme_classic() + ylab("Mean item easiness") + xlab("Language")
#ggsave("xling_diff_by_category_WSprod.pdf", width=11, height=16)
```


## Cross-linguistic similarities 

We look at the Spearman correlation between the item difficulty of each language compared to each other language. 
We might expect this to recapitulate the historical relationship between languages, with more similar languages having more similar item difficulties (e.g., Quebecois and European French).

```{r xling-sim, fig.width=10, fig.height=10, set.cap.width=T, num.cols.cap=2, fig.cap = "Cross-linguistic similarity (Spearman correlation) of IRT item difficulty from the CDI."}
# parm diffs in low- vs. high-data langs
#t.test(xldf$d, xldf_lowd$d) # -.45 vs. -.02 (closer to prior of 0 in low-data langs)
#t.test(xldf$a1, xldf_lowd$a1) # higher discrim in high-data langs

tmp <- get_cross_ling_difficulty_cors(xldf_clean)
prod_cors = tmp$prod_cors
prod_sims = tmp$prod_sims

Colors = brewer.pal(11,"Spectral")
diag(prod_cors) = NA

#pdf(file="xling_difficulty_similarity.pdf", width=13, height=13)
heatmap.2(prod_cors, col=Colors, margins=c(10,10))
#dev.off()

# compare to one of these?
# http://www.elinguistics.net/
# https://www.ezglot.com/most-similar-languages.php
# 28 x 28 = 784 values...scrape?

```


# Cross-validation Results

```{r plot-crossval-fscores, fig.width=7, fig.height=9, fig.cap="Difficulty correlation and overlap by k for the different sublist selection methods.", echo=F}
cv_res_sum <- readRDS(here("data/intermediates/full_vs_swadesh_cv_fscores_allforms.rds"))

cv_output <- cv_res_sum |> 
  group_by(k, sublist) |> 
  summarise(num_overlap = mean(num_overlap, na.rm = TRUE),
            difficulty_cor = mean(difficulty_cor, na.rm = TRUE),
            rmse = mean(rmse, na.rm=T))

best_k <- cv_output |>
  ungroup() |> group_by(sublist) |>
  filter(difficulty_cor == max(difficulty_cor, na.rm = TRUE)) |>
  slice(1) |> ungroup() |> arrange(desc(difficulty_cor)) |>
  relocate(sublist, .before = k) |>
  select(-rmse) |>
  mutate(sublist = ifelse(sublist=="category", "semantic", sublist))

p1 <- ggplot(cv_res_sum,
       aes(x = k, y = num_overlap, col = sublist)) +
  geom_jitter(aes(col = sublist),
              alpha = .1) +
  geom_boxplot(aes(group = interaction(k, sublist))) +
  labs(y = "Overlap") + theme_classic()

p2 <- ggplot(cv_res_sum,
       aes(x = k, y = difficulty_cor, col = sublist)) +
  geom_jitter(aes(col = sublist),
              alpha = .1) +
  geom_boxplot(aes(group = interaction(k, sublist))) +
  labs(y = "Difficulty correlation") + theme_classic()

# full vs. Swadesh fscore correlation for k=23
fscore_cor_t <- t.test(subset(cv_res_sum, sublist=="unstratified" & k==23)$fscore_cor - 
                       subset(cv_res_sum, sublist=="Random" & k==23)$fscore_cor) # n.s.

cv_res_sum$best_k = F
for(i in 1:nrow(best_k)) {
  idx <- which(cv_res_sum$k==best_k[i,]$k & cv_res_sum$sublist==best_k[i,]$sublist)
  cv_res_sum[idx,]$best_k = T
}

p3 <- cv_res_sum |>
  filter(best_k==T) |>
  filter(sublist!="4 strata", sublist!="5 strata") |>
  ggplot(aes(x = language, y = fscore_cor, group = sublist, col = sublist)) +
  geom_point(alpha=.7) + # geom_jitter
  labs(y = "Fscore correlation") +
  theme(legend.position = "bottom") + theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, hjust=0.95, vjust=0.9)) 
# ToDo: maybe sort these by the difference (Swadesh - random) ? 

ggpubr::ggarrange(p1, p2, ncol=1, common.legend = T) # , p3
```

## Swadesh CDI vs. Full CDI:WS

The table below shows the correlation between full CDI and Swadesh-CDI for syntactic category stratification method, for items on at least $k=27$ CDIs, for each of the 32 languages that the IRT models were trained on.

```{r, results='asis'}
#load(here("data/Swadesh_comparisons.Rdata"))
#xx <- readRDS(here("data/intermediates/full_vs_swadesh_fscores_allforms.rds"))
xx <- readRDS(here("data/intermediates/full_vs_swadesh_cv_fscores_allforms.rds"))
#xx_tab <- xtable::xtable(xx, digits=c(3), 
#                       caption = "Swadesh CDI vs. Full CDI scores for 32 training languages.")
#print(xx_tab, type="html", comment = F, table.placement = "H", include.rownames=FALSE)

# ToDo: maybe pivot_wider to show difficulty_cor for the top 3 selection methods?
xx |> filter(k==27, sublist=="syntactic") |>
  rename(`Full vs. S-CDI r` = fscore_cor,
         Language = language,
         Overlap = num_overlap) |>
  select(Language, Overlap, `Full vs. S-CDI r`) |>
  kable(digits=3, caption="Swadesh CDI vs. Full CDI scores for 32 training languages.") |>
  html_table_width(rep(80, ncol(xx)))
```

### Generalization Test

Comparison to the 10 low-data languages [ToDo: with the 10 additional difficult words added]. 
Note that many proposed Swadesh items are not actually on the CDI:WS forms available to test generalization. 

```{r, results='asis'}
# ToDo: get generalization results with 10-item extension
load(here("data/intermediates/generalization_results.rds")) # gen_res, gen_res_sum

gen_xx <- gen_res_sum |> 
  filter(language %in% gen_langs) |>
  group_by(sublist) |>
  rename(Method = sublist) |>
  pivot_wider(id_cols = language, names_from = Method, values_from = sumscore_cor)

gen_xx %>% kable(digits=3, caption="Swadesh CDI vs. Full CDI:WS scores for 10 generalization languages.") %>%
  html_table_width(rep(70, ncol(gen_xx)))
```


## Swadesh-CDI Items

Below we show the full list of 100 Swadesh CDI unilemmas, along with their average cross-linguistic difficulty (d_m), variability in difficulty (d_sd), number of CDI:WS forms they appear on (n, out of 32), semantic category, and lexical category. The semantic and lexical categories are based on American English, as unilemmas sometimes appear in different categories on different forms, as appropriate.

```{r, show-swadesh-list}
load(here("data/intermediates/best_k_swadesh_lists.rds"))
# unstrat_sublist, category_sublist, syntactic_sublist, strat2_sublist

uni_sem_cats <- prod_pars |> group_by(uni_lemma) |>
  count(semantic_category) |> arrange(uni_lemma, desc(n)) |>
  group_by(uni_lemma) |> slice(1)

uni_syn_cats <- prod_pars |> group_by(uni_lemma) |>
  count(syntactic_category) |> arrange(uni_lemma, desc(n)) |>
  group_by(uni_lemma) |> slice(1)

swadesh_sublist <- syntactic_sublist |> select(-category, -count, -mean_prop) |>
  left_join(uni_sem_cats |> select(uni_lemma, semantic_category)) 


en_good_prod <- swadesh_sublist |> 
  left_join(prod_pars |> 
              filter(language=="English (American)") |>  
              select(uni_lemma, semantic_category, syntactic_category)) |> 
  arrange(semantic_category, desc(mean_d))
#kableExtra::kable(en_good_prod, digits=2)
en_good_prod %>% DT::datatable() |> 
  DT::formatRound(columns=3:4, digits=2)
```


## Comparison of Swadesh CDI Concepts to Other Lists

28 of the Swadesh CDI candidates are on the 100-item CDI:WS short (form A), and 26 of the Swadesh candidates are on the WS short (form B).
22 Swadesh CDI concepts are also on the original Swadesh-100 list---and 7 are overlapping with the ASJP (a subset of the Swadesh-100 that performs as well as the Swadesh-100 for purposes of glottochronology).


```{r swadesh-comparison, echo=F}
# https://en.wikipedia.org/wiki/Swadesh_list
# "I","you","we",
swadesh100 = c("1SG","2SG","we","this","that","who","what","not","all","many","one","two","big","long","small",
               "woman","man","person","fish","bird","dog","louse","tree","seed","leaf","root","bark","skin","flesh","blood",
               "bone","grease","egg","horn","tail","feather","hair","head","ear","eye","nose","mouth","tooth","tongue","claw",
               "foot","knee","hand","belly","neck","breasts","heart","liver","drink (action)","eat","bite","see","hear","know","sleep",
               "die","kill","swim","fly","walk","come","lie","sit","stand","give","say","sun","moon","star","water",
               "rain","stone","sand","earth","cloud","smoke","fire","ash","burn","path","mountain","red","green","yellow","white",
               "black","night","hot","cold","full","new","good","round","dry","name")

# https://en.wikipedia.org/wiki/Automated_Similarity_Judgment_Program#Word_list
# ASJP subset of 40 Swadesh items that perform as well as full list
# categories: Body parts, Animals and plants, People, Nature, Verbs and adjectives, Numerals and pronouns
asjp = c("eye","ear","nose","tongue","tooth","hand","knee","blood","bone","breast","liver","skin",
         "louse", # "bug" ?
         "dog","fish (animal)","horn","tree","leaf",
         "person","name", # People; match "name" to "child's own name" / "babysitter's name" / "pet's name" ?
         "sun","star","water (not beverage)","fire","stone","path","mountain","night", # Nature
         "drink (action)","die","see","hear","come", # Verbs and adjectives
         "new","full","one","two","1SG","2SG","we")

#length(intersect(swadesh_sublist$uni_lemma, swadesh100))
#length(intersect(swadesh_sublist$uni_lemma, asjp))


# intersect(swadesh100, subset(xldf, language=="English (American)")$uni_lemma) 
# 55 Swadesh words on English CDI (with "I")
# intersect(swadesh100, xldf$uni_lemma) 
# 83 Swadesh words on full list of unilemmas
#setdiff(swadesh100, unique(xldf$uni_lemma))
# "I" "fly" "fish" - "belly" (belly button?) "louse" "seed" "root" "skin" "flesh" "grease"  "horn"   
# "feather" "claw" "breasts" "liver" "kill" "lie" "water" "ash" "dry" 
```


```{r}
floccia30wordCDI <- c("elephant", "fish (animal)", "lion", "airplane", "car", "ball", "book", "milk", "nose",
                      "shoe", "sock", "bed", "chair", "door", "table", "cup", "scissors", "soap", "flower",
                      "sky", "tree", "mommy", "fall (action)", "play", "big", "cold", "dirty", "hot",
                      "what", "where (question)")

intersect(asjp, floccia30wordCDI)
# "nose"          "fish (animal)" "tree" 

intersect(swadesh100, floccia30wordCDI) # 6
# "what" "big"  "tree" "nose" "hot"  "cold"

intersect(swadesh_sublist$uni_lemma, floccia30wordCDI) # 11
# "what"     "mommy"    "airplane" "milk"     "lion"     "nose"     "elephant" "cold"     "dirty"   "hot" 
```

