---
title: "Appendix to _Measuring Children's Early Vocabulary in Low-Resource Languages Using a Swadesh-style Word List_"
author: "[redacted for anonymous review]"
date: "June 20, 2025"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=F, message=F)

library(png)
library(grid)
library(gt)
library(mirt)
library(tidyverse)
library(ggpubr)
library(tidyboot)
library(here)
library(RColorBrewer)
library(gplots)
library(kableExtra)

html_table_width <- function(kable_output, width){
  width_html <- paste0(paste0('<col width="', width, '">'), collapse = "\n")
  sub("<table>", paste0("<table>\n", width_html), kable_output)
} 

source(here("scripts/04-cross-ling-comparison-helpers.R"))

# production fits
load(here("data/intermediates/xldf_clean_allforms.Rdata")) 
# xldf_clean, prod_pars
# 42 languages, 28541 parameters

xldf_clean <- xldf_clean |> 
  mutate(d = -d)  # easiness -> difficulty

prod_pars <- prod_pars |> 
  mutate(d = -d)  # easiness -> difficulty

load(file=here("data/intermediates/generalization_results.rds"))
# gen_res_sum, gen_res

model_stats <- readRDS(here("data/intermediates/model_stats.rds"))

gen_langs <- model_stats |> filter(participants < 300) |> pull(language)

xldf_lowd <- xldf_clean %>% filter(is.element(language, gen_langs)) # 6324
xldf_train <- xldf_clean %>% filter(!is.element(language, gen_langs)) # 22217


olap <- get_uni_lemma_overlap(xldf_clean)
```

## Unilemma Operationalisation and Updating

Unilemmas are rough cross-linguistic conceptual mappings for words. 
For example, "dog" (English), "perro" (Spanish), and "chien" (French) all seem like they describe the same concept, _dog_. 
We thus have _dog_ as the unilemma for this concept, allowing mapping between different forms in different languages.

Decisions about which words correspond to which unilemmas are necessarily somewhat arbitrary. 
Our design intention for unilemmas was to provide reasonable mappings, lumping together many meanings rather than splitting them, allowing us to build predictive models of cross-linguistic similarities and differences in acquisition.
If we split the conceptual world too finely, it would be difficult to understand how similar the information sources are; if we split too coarsely, we can at least discover that languages differ by the differences in predictors.

However, we also preserve distinctions for concepts that are _prima facie_ similar, but that play very different cultural roles.
For example, although "pan" (Spanish) and "tortilla" (Mexican Spanish[^1]) are relatively similar, they play different cultural roles in Mexican Spanish, and hence we err towards splitting these into separate categories. 

[^1]: Note that "tortilla" in European Spanish refers not to flatbread, but to an egg and potato omelette.

There remained other issues regarding unilemma definitions, including cases of polysemy, handling syntactic categories, and pronouns; in each case, we made decisions based on our best guess of the most basic or common interpretation in the child language context.
Further details can be found in our unilemma policy document, available at [https://github.com/langcog/update_unilemmas/blob/main/unilemma_policy.md](https://github.com/langcog/update_unilemmas/blob/main/unilemma_policy.md).

We then recruited native or advanced proficient speakers of all the languages in Wordbank to help verify and correct unilemmas for the CDI forms. 
Subsequently, two of the authors (GK, AWMT) reviewed the suggestions and made decisions about conflicts, and standardized unilemmas into a canonical form across all instruments.
More details about the recent update can be found at [https://github.com/langcog/update_unilemmas](https://github.com/langcog/update_unilemmas).

# Cross-validation Results

```{r plot-crossval-fscores, fig.width=7, fig.height=9, fig.cap="Difficulty correlation and overlap by k for the different sublist selection methods.", echo=F}
cv_res_sum <- readRDS(here("data/intermediates/full_vs_swadesh_cv_fscores_allforms.rds"))

cv_output <- cv_res_sum |> 
  group_by(k, sublist) |> 
  summarise(num_overlap = mean(num_overlap, na.rm = TRUE),
            difficulty_cor = mean(difficulty_cor, na.rm = TRUE),
            rmse = mean(rmse, na.rm=T))

best_k <- cv_output |>
  ungroup() |> group_by(sublist) |>
  filter(difficulty_cor == max(difficulty_cor, na.rm = TRUE)) |>
  slice(1) |> ungroup() |> arrange(desc(difficulty_cor)) |>
  relocate(sublist, .before = k) |>
  select(-rmse) |>
  mutate(sublist = ifelse(sublist=="category", "semantic", sublist))

p1 <- ggplot(cv_res_sum,
       aes(x = k, y = num_overlap, col = sublist)) +
  geom_jitter(aes(col = sublist),
              alpha = .1) +
  geom_boxplot(aes(group = interaction(k, sublist))) +
  labs(y = "Overlap") + theme_classic()

p2 <- ggplot(cv_res_sum,
       aes(x = k, y = difficulty_cor, col = sublist)) +
  geom_jitter(aes(col = sublist),
              alpha = .1) +
  geom_boxplot(aes(group = interaction(k, sublist))) +
  labs(y = "Difficulty correlation") + theme_classic()

# full vs. Swadesh fscore correlation for k=23
fscore_cor_t <- t.test(subset(cv_res_sum, sublist=="unstratified" & k==23)$fscore_cor - 
                       subset(cv_res_sum, sublist=="Random" & k==23)$fscore_cor) # n.s.

cv_res_sum$best_k = F
for(i in 1:nrow(best_k)) {
  idx <- which(cv_res_sum$k==best_k[i,]$k & cv_res_sum$sublist==best_k[i,]$sublist)
  cv_res_sum[idx,]$best_k = T
}

p3 <- cv_res_sum |>
  filter(best_k==T) |>
  filter(sublist!="4 strata", sublist!="5 strata") |>
  ggplot(aes(x = language, y = fscore_cor, group = sublist, col = sublist)) +
  geom_point(alpha=.7) + # geom_jitter
  labs(y = "Fscore correlation") +
  theme(legend.position = "bottom") + theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, hjust=0.95, vjust=0.9)) 
# ToDo: maybe sort these by the difference (Swadesh - random) ? 

ggpubr::ggarrange(p1, p2, ncol=1, common.legend = T) # , p3
```

Figure \@ref(fig:plot-crossval-fscores) shows the full cross-validation results for the different sublist selection methods, showing the number of overlapping items and item-wise correlations in difficulty as a function of $k$, the minimum number of languages in which the items appear.

## Swadesh CDI vs. Full CDI:WS

Table \@ref(tbl:full-correl) shows the correlation between full CDI and Swadesh-CDI for syntactic category stratification method, for items on at least $k=27$ CDIs, for each of the 32 languages that the IRT models were trained on.

```{r full-correl, results='asis'}
#load(here("data/Swadesh_comparisons.Rdata"))
#xx <- readRDS(here("data/intermediates/full_vs_swadesh_fscores_allforms.rds"))
xx <- readRDS(here("data/intermediates/full_vs_swadesh_cv_fscores_allforms.rds"))
#xx_tab <- xtable::xtable(xx, digits=c(3), 
#                       caption = "Swadesh CDI vs. Full CDI scores for 32 training languages.")
#print(xx_tab, type="html", comment = F, table.placement = "H", include.rownames=FALSE)

# ToDo: maybe pivot_wider to show difficulty_cor for the top 3 selection methods?
xx |> filter(k==27, sublist=="syntactic") |>
  rename(`Full vs. S-CDI r` = fscore_cor,
         Language = language,
         Overlap = num_overlap) |>
  select(Language, Overlap, `Full vs. S-CDI r`) |>
  kable(digits=3, caption="Swadesh CDI vs. Full CDI scores for 32 training languages.") |>
  html_table_width(rep(80, ncol(xx)))
```

<!-- ## Swadesh-CDI Items -->

<!-- Below we show the full list of 100 Swadesh CDI unilemmas, along with their average cross-linguistic difficulty (d_m), variability in difficulty (d_sd), number of CDI:WS forms they appear on (n, out of 32), semantic category, and lexical category. The semantic and lexical categories are based on American English, as unilemmas sometimes appear in different categories on different forms, as appropriate. -->

```{r show-swadesh-list, eval=F}
load(here("data/intermediates/best_k_swadesh_lists.rds"))
# unstrat_sublist, category_sublist, syntactic_sublist, strat2_sublist

uni_sem_cats <- prod_pars |> group_by(uni_lemma) |>
  count(semantic_category) |> arrange(uni_lemma, desc(n)) |>
  group_by(uni_lemma) |> slice(1)

uni_syn_cats <- prod_pars |> group_by(uni_lemma) |>
  count(syntactic_category) |> arrange(uni_lemma, desc(n)) |>
  group_by(uni_lemma) |> slice(1)

swadesh_sublist <- syntactic_sublist |> select(-category, -count, -mean_prop) |>
  left_join(uni_sem_cats |> select(uni_lemma, semantic_category)) 


en_good_prod <- swadesh_sublist |> 
  left_join(prod_pars |> 
              filter(language=="English (American)") |>  
              select(uni_lemma, semantic_category, syntactic_category)) |> 
  arrange(semantic_category, desc(mean_d))
#kableExtra::kable(en_good_prod, digits=2)
en_good_prod %>% DT::datatable() |> 
  DT::formatRound(columns=3:4, digits=2)
```


## Comparison of Swadesh CDI Items to Other Lists

There are several short vocabulary lists of 100 items that have been constructed for different purposes, which serve as useful comparisons to the Swadesh CDI list.
Here, we list the items on the S-CDI list, and whether they also occur in other short vocabulary lists---namely, lists used in lexicostatistics (the original Swadesh-100 list CITE, the Leipzig--Jakarta list CITE), short form CDI lists used for vocabulary assessment (the American English CDI:WS Short Forms A and B CITE, and the Oxford Short Form CDI), and lists used for pedagogical instruction (the Fry-100 Sight Words list CITE, and the Children's Picture Book Sight Words list CITE).
These lists range in overlap with the S-CDI, from 13 items (Fry-100 and Children's Picture Book) to 25 items (Oxford Short Form CDI); the overlap in items suggests that the S-CDI list captures some of the relevant dimensions of vocabulary (consistency across languages, utility in assessment, and pedagogical importance).

```{r form-compare, tab.cap="Comparison of the S-CDI list with other 100-item vocabulary lists."}
form_comparison <- read_csv(here("data", "form_comparison.csv"))

form_comparison |> 
  bind_rows(
    form_comparison |> 
      summarise(across(where(is.numeric), 
                       list(`S-CDI overlap` = \(x) sum(x & !str_starts(`S-CDI items`, "_"), na.rm = TRUE),
                            `S-CDI-E overlap` = \(x) sum(x, na.rm = TRUE)),
                       .names = "{.col}.{.fn}")) |> 
      pivot_longer(cols = everything()) |> 
      separate_wider_delim(cols = name, delim = ".", names = c("form", "Semantic category")) |> 
      pivot_wider(names_from = form, values_from = value) |> 
      mutate(`S-CDI items` = "")
  ) |> 
  mutate(across(everything(), 
                \(x) x |> as.character() |> replace_na("") |> str_replace("^1$", "✓")),
         `Semantic category` = ifelse(`Semantic category` != lag(`Semantic category`) | 
                                        is.na(lag(`Semantic category`)), 
                                      `Semantic category`, "")) |> 
  gt() |> 
  tab_spanner_delim("_") |> 
  fmt_markdown(columns = everything()) |> 
  cols_align(align = "center", columns = -c(`Semantic category`, `S-CDI items`)) |> 
  # cols_merge(
  #   columns = c(`Semantic category`, `S-CDI items`),
  #   pattern = "{1}",
  #   rows = c(1:)
  # ) |>
  tab_style(
    style = cell_borders(
      sides = "top",
      color = "black",
      weight = px(1.5),
      style = "solid"
    ),
    locations = cells_body(
      rows = `Semantic category` == "S-CDI overlap"
    )
  )
```

